HO unification from object language to meta language

### 2024-03-18 paper/main.tex
2

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\solve}{\ensuremath{\mathrm{check}}}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[5]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto #5}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \solve(#6,#3 \cup #4) \mapsto #5}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3}{\sigma_{p}}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
$$
\begin{array}{l}
\stepF{\mathcal{S}}{p}{\rho}{\rho''}
\deff
\stepFD{\mathcal{S}}{p}{\rho}{\rho'}{\rho'}\\
\runF{\mathcal{S}}{\mathcal{N}}{\rho}
\deff
\runFD{\mathcal{S}}{\mathcal{N}}
\end{array}
$$

We simulate each run in \Fo with a run in \Ho as follows.
Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma''} \deff\\
  \qquad\stepHD{\mathcal{T}}{p}{\sigma}{\sigma'}{\sigma''}{\mathbb{L}}\\
\runH{\mathcal{S}}{\mathcal{N}}{\rho} \deff \\
  \qquad \mathcal{T} \times \mathbb{M} \times \mathbb{L} = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\\
  \qquad \runHD{\mathcal{T}}{\mathcal{N}}{\mathbb{L}}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mathbb{M}}{\mathbb{L}}{\rho_{\mathcal{N}}}
\end{array}
$$

\noindent
Here \hstep is made of two sub-steps: a call to \Ue (on the compiled
terms) and a call to \solve on the set of links. We claim the following:

@@

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of~ \hrun, we have that
$\forall p \in 1 \ldots \mathcal{N}$
$$
\stepF{\mathcal{S}}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\mathcal{T}}{p}{\sigma_{p-1}}{\mathbb{L}}{\sigma_p}
$$
\end{proposition}
\noindent
Note that this property guarantees that a failure in the \Fo run
is matched by a failure
in \Ho
\emph{at the same step}.

\newcommand{\maybebeta}{\ensuremath{\beta}\xspace}
\begin{definition}[\maybebeta]
  $t$ is in \maybebeta if it contains a subterm matching the
  pattern \elpiIn{app[uva _ _|_]}
\end{definition}

\newcommand{\maybeeta}{\ensuremath{\eta}\xspace}
\begin{definition}[\maybeeta]
  $t$ is in \maybeeta if it contains a subterm matching the
  pattern \elpiIn{lam x\ .. lam y\ app[con _,x, ...,uva _ [y]]}\\
   or
  \elpiIn{lam x\lam y\uva _ [y,x]}
\end{definition}

\begin{definition}[Normal form]
  We write $norm(\mathcal{T})$ when $\mathcal{T} \cap (\maybebeta \cup \maybeeta) = \emptyset$
\end{definition}

We write $\sigma\mathcal{T} = \{ \sigma t_i | t_i \in \mathcal{T}\}$.

\begin{proposition}[Normal form preservation]\label{prop:nf}
  $norm(\sigma\mathcal{T}) \land \mathrm{step'}(p,\sigma) \mapsto \sigma'
  \Rightarrow norm(\sigma\mathcal{T})$   
\end{proposition}

If we focus on a single unification problem, that is
$\mathcal{S} = \{ s_1, s_2 \}$ we can define \Uo as follows:
$$
\begin{array}{l}
s_1 \Uo s_2 \mapsto \rho \deff\\
\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\\
\quad    t_1 \Ue t_2 \mapsto \sigma' \land
    \solve~(\{l_1,l_2\},\sigma') \mapsto \sigma'' \land\\
\quad \D{\sigma''}{\{m_1,m_2\}}{\{l_1,l_2\}}{\rho}
\end{array}
$$
\noindent
Note that proposition \ref{prop:nf} does not hold in this case since
decompilation can introduce terms in \maybebeta or \maybebeta. Nevertheless
\Uo has the following properties (where we
write $s \in \llambda$ if all unif variables in
$s$ are applied to distinct bound variables):

@@

Property \ref{prop:simulation} states that \Uo is resilient to problems outside
\llambda if a third party provides a (partial) solution for its problem. Since
we are interested in using \Uo in a proof search procedure, made of a sequence
of unification problems, not necessarily in \llambda. In practice it is often
the case the order in which these problems are stated matters. A Typical example
is the following problem

\begin{elpicode}
app [F, con"a"] = app[con"f", con"a", con"a"]
\end{elpicode}

\noindent
preceded by

\begin{elpicode}
F = lam x\ app[con"f",x,x]
\end{elpicode}

\noindent
becomes solvable in DTT trivially, since the term is ground (hence in \llambda),
but if one substitutes \elpiIn{F} in the LHS does not find,
structurally, the RHS
hence \Ue would fail (since \Ue does not know about the $\beta$ rule of DTT).
Our compiler takes care of making property \ref{prop:simulation} hold, see
section XXX.

Property \ref{prop:fidelity} is also relevant to use \Uo for logic programming.
In particular we want failures to occur as early as possible, so want
the check phase to take place immediately after \Ue, and fail if need be.

@@

@@

\section{Alternative encodings and related work}

@@

\begin{elpicode}
type deref subst -> tm -> tm -> o.                   ~($\customlabel{pred:deref}{\sigma t}$)~
deref S (app A) (app B) :- map (deref S) A B.
deref S (lam F) (lam G) :- 
  pi x\ deref S x x => deref S (F x) (G x).
deref _ (con C) (con C).
deref S (uva N A) R :- set? N S F, move F A T, deref S T R.
deref S (uva N A) (uva N B) :- unset? N S, map (deref S) A B.
type move assignment -> list tm -> tm -> o.
move (abs F) [X | L] R :- move (F X) L R.
move (val A)         [] A :- !.
move (val (app A))   L (app X) :- append A L X.
move (val (uva N A)) L (uva N X) :- append A L X.
move (val (con H))   L (app [con H | L]).
\end{elpicode}

>>>>

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\solve}{\ensuremath{\mathrm{check}}}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[5]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto #5}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \solve(#6,#3 \cup #4) \mapsto #5}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3}{\sigma_{p}}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
$$
\begin{array}{l}
\stepF{\mathcal{S}}{p}{\rho}{\rho''}
\deff
\stepFD{\mathcal{S}}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
\runF{\mathcal{S}}{\mathcal{N}}{\rho}
\deff
\runFD{\mathcal{S}}{\mathcal{N}}
\end{array}
$$

We simulate each run in \Fo{} with a run in \Ho as follows.
Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma''} \deff\vspace{2pt}\\
  \qquad\stepHD{\mathcal{T}}{p}{\sigma}{\sigma'}{\sigma''}{\mathbb{L}}\vspace{2pt}\\
\runH{\mathcal{S}}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \mathcal{T} \times \mathbb{M} \times \mathbb{L} = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
  \qquad \runHD{\mathcal{T}}{\mathcal{N}}{\mathbb{L}}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mathbb{M}}{\mathbb{L}}{\rho_{\mathcal{N}}}
\end{array}
$$

\noindent
Here \hstep{} is made of two sub-steps: a call to \Ue (on the compiled
terms) and a call to \solve{} on the set of links. We claim the following:

@@

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of~ \hrun, we have that
$\forall p \in 1 \ldots \mathcal{N}$
$$
\stepF{\mathcal{S}}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\mathcal{T}}{p}{\sigma_{p-1}}{\mathbb{L}}{\sigma_p}
$$
\end{proposition}
\noindent
In particular this property guarantees that a \emph{failure} in the \Fo{} run
is matched by a failure in \Ho{} \emph{at the same step}. We consider this
property very important from a practical point of view since it guarantees
that the execution traces are strongly related and in turn this enables a user
to debug a logic program in \Fo{} by looking at its execution trace in
\Ho{}.

We can define $s_1 \Uo{} s_2$ by specializing the code of \hrun{} to
$\mathcal{S} = \{ s_1, s_2 \}$ as follows:
%
$$
\begin{array}{l}
s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
\quad    t_1 \Ue t_2 \mapsto \sigma' \land
    \solve~(\{l_1,l_2\},\sigma') \mapsto \sigma'' \land\vspace{2pt}\\
\quad \D{\sigma''}{\{m_1,m_2\}}{\{l_1,l_2\}}{\rho}
\end{array}
$$

@@

Property \ref{prop:simulation} states that \Uo, hence our compilation scheme,
is resilient to unification problems outside \llambda solved by
a third party. We believe this property is of practical interest since we
want the user to be able to add heuristics via hand written rules
to the ones obtained by our compilation scheme. A Typical example
is the following problem \ref{problem:q} that is outside \llambda:

\begin{elpicode}
app [F, con"a"] = app[con"f", con"a", con"a"]   ~\customlabel{problem:q}{(q)}~
F = lam x\ app[con"f",x,x]                      ~\customlabel{heuristic:h}{(h)}~
\end{elpicode}

\noindent
Instead of rejecting it our scheme accepts it and guarantees that if
\ref{heuristic:h} is given (after the compilation part of the scheme, as
a run time hint) then ...

@@

\newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
\newcommand{\maybebeta}{\ensuremath{\Diamond\beta}\xspace}
\begin{definition}[\maybeeta]
  $\maybeeta = \{ t | \exists \rho, \rho t ~\mathrm{is~an~eta~expansion} \}$
\end{definition}

\noindent
An example of term $t$ in \maybeeta{} is
$\lambda x.\lambda y.F~y~x$
since the substitution
$\rho = \{ F \mapsto \lambda a.\lambda b.fba\}$
makes $\rho t = \lambda x.\lambda y.f x y$
that is the eta long form of $f$.

\begin{definition}[\maybebeta]
  $t$ is in \maybebeta if it contains a subterm matching the
  pattern \elpiIn{app[uva _ _|_]}
\end{definition}

\newcommand{\normal}{\ensuremath{\mathrm{normal}}}
\begin{definition}[Normal form]
Given a set of terms $X \subseteq \Ho{}$,
$$
\normal(X) =
X \cap (\maybebeta{} ~\cup~ \maybeeta{}) = \emptyset
$$
\end{definition}

\noindent
We write $\sigma X = \{ \sigma t_i | t_i \in X\}$.

\begin{proposition}[Normal form preservation]\label{prop:nf}
$\forall \mathcal{T}, \forall \mathbb{L}, \forall p,$
$$
\normal(\sigma\mathcal{T}) \land
\stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma'}
\Rightarrow \normal(\sigma'\mathcal{T})
$$
\end{proposition}

@@

Note that proposition \ref{prop:nf} does not hold in this case since
decompilation can introduce terms in \maybebeta or \maybebeta.
\section{Alternative encodings and related work}

@@

\begin{elpicode}
type deref subst -> tm -> tm -> o.                   ~($\customlabel{pred:deref}{\sigma t}$)~
deref S (app A) (app B) :- map (deref S) A B.
deref S (lam F) (lam G) :- 
  pi x\ deref S x x => deref S (F x) (G x).
deref _ (con C) (con C).
deref S (uva N A) R :- set? N S F, move F A T, deref S T R.
deref S (uva N A) (uva N B) :- unset? N S, map (deref S) A B.
type move assignment -> list tm -> tm -> o.
move (abs Bo)        [H|L] R :- move (Bo H) L R.
move (val A)         []    A :- !.
move (val (uva N A)) L     (uva N X) :- std.append A L X.
\end{elpicode}
